The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}

### 2017-05-31 GRB111117A.tex
TK: Typographical changes, nothing on the science

\title{The host galaxy of the short GRB~111117A at z = 2.211\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}}

@@

\abstract{
Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to
localize. We here present a spectroscopic observation of the host galaxy of GRB
111117A and measure its redshift to be $z = 2.211$, making it the most distant
high-confidence short GRB with a secure redshift measurement. This value
supersedes a previously estimated, lower redshift value. Spectroscopic
observations along with a wealth of imaging data allow us to place tight
constraints on the nature of the host galaxy at an unprecedented distance.

The rest frame X-ray derived hydrogen column density is high compared to a
complete sample of sGRBs and seems to follow the evolution with redshift as
traced by the hosts of long GRBs (lGRBs). This could indicate that the part of
the sGRB population hosted in late-type galaxies live in similar environments to
that of lGRB hosts.

The host lies in the brighter end of the expected host brightness distribution
at $z = 2.211$, and is actively star forming. Using the host as a benchmark for
redshift determination, we assess that only 30 per cent of sGRB redshifts should
be missed due to host faintness at this redshift. This indicates that we are not
missing redshifts for a dominant fraction of the sGRB hosts at this redshift.
The redshift of GRB~111117A is evidence against a lognormal delay-time model for
sGRBs through the predicted redshift distribution of sGRBs, which is very
sensitive to high-z sGRBs.

From the age of the universe at explosion time, an initial progenitor separation
of a$_0 < 3.2 $R$_\odot$ is required for the case of a binary neutron star (NS)
system. This puts constraints on the progenitor system evolution up to the time
of explosion.
}

@@

There is now mounting evidence that most short-duration gamma-ray bursts (GRBs)
come from the merger of neutron stars (NSs), either with another NS, or a black
hole, due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The absence of associated
supernovae in deep searches \citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b}
supports this idea and distinguishes the physical origin of sGRBs from their
long-duration counterparts. %\citep[albeit see also][]{Fynbo2006b, Valle2006,
% Gal-Yam2006}.
The range of host galaxy types, in some cases apparently absent, and typically
more massive and on average less actively star-forming than long GRB hosts 
\citep{Fong2013b}, as well as their positions within hosts \citep{Fong2013a},
suggest an origin for the progenitors that can be very long-lived, and
associated with stellar mass rather than purely star-formation rate
\citep{Berger2014}. The bulk of sGRBs have associated galaxies at relatively low
redshifts with a median $z\sim0.5$ \citep{Berger2014} and because most redshifts
are measured from the hosts, this value is biased towards low $z$.

@@

The total lifetimes of NS binaries is dependent on their initial separations and
subsequent inspiral times, and impacts the timing and distribution of the
enrichment of the ISM and subsequent stars and planets with heavy $r$-process
elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some limits can be
calculated based on models of star-formation histories of, and the spatial
distribution of sGRBs in, their host galaxies \citep[][]{Berger2014}. The most
distant cosmological bursts, however, offer direct, hard limits.

In this \emph{Letter} we present the spectrum of the host galaxy of the short
GRB 111117A at z=2.211, a significantly higher redshift than the previously
estimated redshift based on photometric studies. We present the rest frame burst
properties based on this new distance compared to previous analyses
\citep{Margutti2012,Sakamoto2013} and revisit the host properties derived from
the new solution to the SED fit. %While no optical afterglow was detected, the
% excellent localization from a detection
%of the X-ray afterglow by the \emph{Chandra X-ray Observatory} also allows
%us to discuss the positioning and environmental properties of this remarkably
%distant short GRB.

Throughout this \emph{Letter} we use the $\Lambda$CDM cosmology provided by
\citet{Planck2015} in which the universe is flat with $H_0 = 67.7$ km s$^{-1}$
and $\Omega_m = 0.307$ and report all magnitudes in the AB system.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{Imaging of the field with the X-shooter slit overlaid. Only one slit is shown, despite 4 epochs of spectroscopic observations because of the similarity in position angle. The image is the FORS2 R-band image and%, for which the photometry is shown in Fig. \ref{fig:SED}. 
	the blue asterisk i ndicates the GRB position as derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. The errorcircle of the \emph{Chandra}-position is too small to resolve in the image.}
	\label{fig:spec_setup}
\end{figure}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, in four seperate epochs. The
burst was followed up 38 hours after the Burst Alert Telescope (BAT) trigger
under ESO programme 088.A-0051 (PI: Fynbo) and again later using a different ESO
program 091.D-0904 (PI: Hjorth). X-shooter can cover the wavelength range from
3000 \AA~to 24800 \AA~(21000 \AA~when the K-band blocking filter is used)~across
three spectroscopic arms, simultaneously illuminated through the use of
dichroics. The bias-correction, flat-fielding, order tracing, wavelength
calibration, rectification, and flux calibration is then carried out using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010}. 
%Because the spectra are curved across each detector, a rectification algorithm
%is employed which introduces correlations between neighboring pixels. We select
%a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the
%degree of correlation while conserving the maximal resolution.
The observations are combined and extracted using scripts described in Selsing
et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The
signal-to-noise of the continuum in the near-infrared arm is too low to use the
optimal extraction algorithm \citep{Horne1986} and therefore the extraction is
carried out with a simple aperture.

An overview of the spectroscopic observations is given in Tab. \ref{tab:spec_overview}, and the position of the slit on the target is shown in Fig. \ref{fig:spec_setup}.

@@

We determine a redshift of $z = 2.211$ from the simultaneous detection of
emission lines interpreted as \lya, \oii, \hb, \oiii$\lambda$5007, and \ha, with
\hb~detected at low significance ($\sim 3 \sigma$). \ha~is only visible in the
first epoch, due to the K-band blocking filter used for the remainder
observations. The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and
show significant velocity structure along the slit. A drop in the continuum to
the blue of the \lya~line further supports the inferred redshift.

@@

Using the integrated flux of \ha, we can infer the SFR \citep{Kennicutt1998}.
From the SED-fit, see Sect. \ref{SED}, and the detection of \lya, the host is
constrained to contain very little or no dust which is also confirmed by the
shape of the spectral continuum, which is why we do not apply a dust-correction.
Because of the redshift, \ha~falls on the far end of the NIR-arm where the night
sky is very bright and additionally, several bright sky-lines intersect the
line, making an accurate estimate of the \ha-flux difficult. A limit on the SFR
is obtained by integrating the part of \ha~free of contamination and
conservatively correcting for the missing fraction by assessing the fraction
integrated. Converting to a \citet{Chabrier2003} initial mass function using
\citet{Madau2014} we get a limit of SFR > 7 M$_\odot$ yr$^{-1}$.

@@

We additionally imaged the field of GRB 111117A in multiple broad-band filters
using the VLT equipped with FORS2 ($gRIz$ filters) and HAWK-I
($JHK_{\mathrm{s}}$ filters), long after the burst has faded. These new data are
complemented by a re-analysis of some of the imagery used in
\citet{Margutti2012} and \citet{Sakamoto2013} that are available to us here (GTC
$gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in Tab.~\ref{tab:phot_overview}.

All data were reduced, analyzed and fitted in a similar manner as we have done
previously and described in detail in \citet{Kruhler2011a} and, more recently,
in \citet{Schulze2016}. Very briefly, we use our own python and IRAF routines to
perform a consistent standard reduction of the available imaging which includes
bias/flat-field correction, de-fringing (if necessary), sky-subtraction, and
stacking of individual images for a given instrument in a given filter. The
photometry of the host was tied against magnitudes of field stars from the SDSS
and 2MASS catalogs in the case of $grizJHK_{\mathrm{s}}$ filters. For our $R$
and $I$-band photometry we used the color transformations of
Lupton\footnote{\url{https://www.sdss3.org/dr8/algorithms/sdssUBVRITransform.php}}. We subsequently convert all magnitudes into the AB system if necessary, and correct for the Galactic foreground of $E_{B-V}=0.027~\mathrm{mag}$.

The multi-color spectral energy distribution (SED) is then fit by
\citet{Bruzual2003} stellar population synthesis models based on a
\citet{Chabrier2003} initial mass function in \emph{LePhare} \citep{Ilbert2006},
where the redshift is fixed to the spectroscopic value of $z=2.211$. The best
model is obtained with an unreddened galaxy template, and returns physical
parameters of luminosity ($M_B=-22.0\pm0.1$\,mag), stellar mass
($\log(M_{\star}/M_\odot) = 9.9\pm0.2$), single stellar population age ($\tau =
0.5_{-0.25}^{+0.5}$ Gyr) and star-formation rate
($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). %We show the SED
% fit in Fig. \ref{fig:SED}.

Noteworthy is the discrepancy of both our new VLT/FORS2 photometry and the
re-analysis of Gemini data to the $z$-band data of \citet{Margutti2012} and
\citet{Sakamoto2013}. Both report $z$-band photometry brighter by 0.8-1.0 mag
than what we derive here, whereas bluer data are in excellent agreement with
previously published values.

In fact, the large $i-z$ color is mistakenly interpreted as a 4000\,\AA\,break
driving the galaxy photometric redshift of both of these earlier works. Using
the revised photometry from Table~\ref{tab:phot_overview}, the photo-$z$ of the
galaxy is $z_{\mathrm{phot}}=2.04_{-0.21}^{+0.19}$, consistent with the
spectroscopic value at the 1~$\sigma$ confidence level.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{XRT temporal and spectral analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We retrieved the automated data products provided by the \textit{Swift}-XRT GRB
repository\footnote{\url{http://www.swift.ac.uk/xrt\_products/00507901.}
	\todo{Broken link, Paolo?}} \citep{Evans2009}. The X-ray afterglow light curve
can be fit with a single power-law decay with index
$\alpha=1.27_{-0.10}^{+0.12}$. We performed a time-integrated spectral analysis
using data in photon counting (PC) mode in the widest time epoch where the
$0.3-1.5\,\mathrm{keV}$ to $1.5-10\,\mathrm{keV}$ hardness ratio is constant
(namely, from $t-T_0 = 205$ s to $t-T_0 = 203.5$ ks, for a total of 29.1 ks of
data), in order to prevent spectral changes that can affect the X-ray column
density determination. The obtained spectrum is well described by an absorbed
power-law model. The best-fit spectral parameters are a photon index of $2.1 \pm
0.4$ and an intrinsic $N_{H}$ of $2.4_{-1.6}^{+2.4} \times 10^{22}$ cm$^{-2}$
($z=2.211$), assuming the Galactic $N_{H}$ in the burst direction of $4.1 \times
10^{20}$ cm$^{-2}$.

@@

As already pointed out \citep{Margutti2012, Sakamoto2013}, GRB~111117A very
securely belongs the short class of GRBs. Because the observed classification
indicators, $T_{90}$ and hardness ratio, do not depend strongly on redshift
\citep{Littlejohns2013a}, the updated redshift does not change this designation
significantly. The intrinsic luminosity increases, as visible in the light curve
in Fig. \ref{fig:sxray_lightcurve}, but it still is sub-luminous compared to
majority of long GRBs. \citet{Bromberg2013} investigated the degree to which the
long and short population distributions overlap and quantified the certainty in
the class membership. GRB~111117A has $96_{-5}^{+3}$ percent probability of
being short. Compared to the other two highest-z short candidates, GRB060121
\citep{DeUgartePostigo2006, Levan2006} at $1.7 \lesssim z \lesssim 4.5$
($17_{-15}^{+14}$ per cent) and GRB090426 \citep{Antonelli2009, Levesque2010,
	Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$ per cent), the certainty in class
membership for GRB~111117A is much higher.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/XLC_111117A_rest.pdf}
	\caption{Restframe XRT lightcurve of GRB~111117A, compared to the general population of XRT lightcurves of GRBs. The grey shared region is a compilation of long GRB lightcurves where the color represents density and the light blue is other short GRB lightcurves for which redshifts has been determined. Despite the remarkably high redshift, the burst’s luminosity is comparable to the bulk of the short burst population and still subluminous compared to the lGRB population.}
	\label{fig:sxray_lightcurve}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Restframe $N_H$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We plot the recalculated $N_H$ in Fig. \ref{fig:NH_z} where we compare with the
distributions of complete samples of both long and short GRBs. The long sample
is from \citet{Arcodia2016} and the short sample is from \citet{DAvanzo2014a}.
17 of the 99 long bursts do not have redshifts and likewise for 5 out of 16 for
the short sample. Bursts without redshifts have been excluded for both groups.
GRB~111117A occupies a unique position in Fig. \ref{fig:NH_z} with the highest
$N_H$ of all short burst. The short sample, excluding GRB~111117A, is located at
low redshifts(z < 1) and are found to populate a similar column density
environment to long GRBs at similar redshifts \citep{DAvanzo2014a}. The inferred
hydrogen column for GRB~111117A seems to follow the trend with increasing $N_H$
as a function of redshift as found for the long GRB hosts \citep{Arcodia2016}.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_z.pdf}
	\caption{Rest frame, X-ray derived hydrogen column densities for GRB~111117A compared to complete samples of both long and short populations. In red is shown the long sample of \citet{Arcodia2016}. The detections are replaced with contours for clarity and the limits are shown with arrows. In blue is the sample from \citet{DAvanzo2014a}. GRB~111117A is an outlier from the short sample, but seems to follow the distribution of long GRBs.}
	\label{fig:NH_z}
\end{figure}

@@

%From the clear host association, GRB~111117A does not belong to the hostless class of GRBs \citep{Berger2010a} and because the host exhibits clear emission lines indicative of a population of relatively young stars, 
As the majority of short GRBs \citep{Fong2013b}, the host of GRB~111117A is a
late-type galaxy and is entirely consistent in terms of stellar mass and stellar
age with the general population of short GRB hosts($\left\langle M _*
\right\rangle = 10^{10.1} M_{\odot}$ and $\left\langle \tau _* \right\rangle =
0.3 $ Gyr) \citet{Leibler2010}. The SFR, on the other hand, is $\sim$1 order of
magnitude higher than the typical SFR for short GRB hosts galaxies
\citep{Berger2014} and more similar to the SFRs found in the hosts of long GRBs
at a corresponding redshift \citep{Kruhler2015}. Only two hosts in the sample of
short GRBs compiled in \citet{Berger2014} have a more vigorous star formation,
meaning that it is in the very upper end of the star formation distribution,
while not a complete outlier. This is likely a selection effect, because a less
star-forming galaxy would exhibit weaker emission lines, thus making the
redshift harder to determine. Being a late-type host, both the stellar mass and
sSFR are entirely consistent with the expected host of
sGRBs\citep{Behroozi2014}.

That we detect \lya~is consistent with the SED-inferred absence of dust, despite
the moderate stellar age and the high X-ray derived hydrogen column density
which would suggest the opposite. The centroid of the Ly$\alpha$ emission is
found to be redshifted by $\sim$ 220 km s$^{-1}$ with respect to systemic, which
is similar to what is found for long GRB hosts \citep{Milvang-Jensen2012a} where
the outflow is attributed to star formation.

>>>>

\title{The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}}

@@

% TK: I made the following typographical changes:
% All physical constants are in italics, all mathematical constants are normal, so SFR
% used in formulas is italics, while it is not italics when used in text as an acronym
% Super- and subscripts are usually normal
% Write out all numbers < 12
% Added fixed spaces in GRB, e.g., GRB~111117A and units (this prevents a line break)
% Thin spaces \, between thousands when > 9999
% Table is never abbreviated
% Tried to make the text a bit easier to read, e.g., things like the following:
% Conclusions: Part of the reason is attributed ... -> The reason is ...

\abstract{
% TK I rephased the abstract a bit, check if you like it or not. The original version is commented out
It is notoriously difficult to localize short $\gamma$-ray bursts (sGRBs) and their hosts to measure their redshifts. These measurements, however, are critical to constrain sGRB progenitors 
and delay time models. Here, 
%Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to
%localize. 
we present spectroscopy of the host galaxy of GRB~111117A 
and measure its redshift to be $z = 2.211$. This makes GRB~111117A the most distant
high-confidence short GRB detected to date.
%with a secure redshift measurement. 
Our spectroscopic redshift supersedes a lower redshift previously estimated from photometry. %Spectroscopic observations along with a wealth of imaging data allow us to place tight
%constraints on the nature of the host galaxy at an unprecedented distance.
We use the spectroscopic redshift, as well as new imaging data to constrain the nature of the host galaxy as well as physical parameters of the GRB. The rest frame X-ray derived hydrogen column density, for example, is high compared to a complete sample of sGRBs and seems to follow the evolution with redshift as traced by the hosts of long GRBs (lGRBs). This could indicate that the part of the sGRB population hosted in late-type galaxies live in similar environments to
that of lGRB hosts.

The host lies in the brighter end of the expected host brightness distribution
at $z = 2.211$, and is actively forming stars. Using the host as a benchmark for
redshift determination, we assess that only 30 per cent of sGRB redshifts should
be missed due to host faintness at $z\sim2$. This indicates that we are not
missing redshifts for a dominant fraction of the sGRB hosts in this redshift range.
The high redshift of GRB~111117A is evidence against a lognormal delay-time model for
sGRBs through the predicted redshift distribution of sGRBs, which is very
sensitive to high-$z$ sGRBs.

From the age of the universe at explosion time, an initial progenitor separation
of $a_0 < 3.2~R_\odot$ is required in the case of a binary neutron star system. 
This puts constraints on the progenitor system evolution up to the time
of explosion.
}

@@

There is now mounting evidence that most short-duration gamma-ray bursts (GRBs)
come from the merger of neutron stars (NSs), either with another NS, or a black
hole, due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The absence of associated
supernovae in deep searches \citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b}
supports this idea and distinguishes the physical origin of sGRBs from their
long-duration counterparts. %\citep[albeit see also][]{Fynbo2006b, Valle2006,
% Gal-Yam2006}.

% TK I rephrased this paragraph, i found the sentences too long and split . Original
% version is commented out.

The host galaxies of short GRBs are diverse. They are more massive and less actively 
star-forming on average than long GRB hosts \citep{Fong2013b}, while in some cases, no host 
galaxy can be identified. Together with their position 
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can be very
long-lived, and is associated with stellar mass rather than star-formation rate
\citep{Berger2014}. The median redshift for sGRBs is $z\sim0.5$ \citep{Berger2014}, and
because most of these measurements come from the associated hosts, it is arguably biased
towards lower redshifts.

%The range of host galaxy types, in some cases apparently absent, and typically
%more massive and on average less actively star-forming than long GRB hosts 
%\citep{Fong2013b}, as well as their positions within hosts \citep{Fong2013a},
%suggest an origin for the progenitors that can be very long-lived, and
%associated with stellar mass rather than purely star-formation rate
%\citep{Berger2014}. The bulk of sGRBs have associated galaxies at relatively low
%redshifts with a median $z\sim0.5$ \citep{Berger2014} and because most redshifts
%are measured from the hosts, this value is biased towards low $z$.

@@

The total lifetimes of NS binaries depends on their initial separations and
subsequent inspiral times, and impacts the timing and distribution of the
enrichment of the ISM and subsequent stars and planets with heavy $r$-process
elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some limits can be
calculated based on models of star-formation histories of, and the spatial
distribution of sGRBs in, their host galaxies \citep[][]{Berger2014}. The most
distant cosmological bursts, however, offer direct, hard limits.

In this \emph{Letter} we present the spectrum of the host galaxy of the short GRB~111117A ($T_{90}=0.46$~s) and measure its redshift to be $z=2.211$. This value is significantly higher 
than the previously estimated redshift based on photometric studies. 
We present the GRB's rest frame 
properties based on this new distance compared to previous analyses
\citep{Margutti2012,Sakamoto2013} and revisit the host properties derived from
the new solution to the SED fit. %While no optical afterglow was detected, the
% excellent localization from a detection
%of the X-ray afterglow by the \emph{Chandra X-ray Observatory} also allows
%us to discuss the positioning and environmental properties of this remarkably
%distant short GRB.

Throughout this \emph{Letter} we use the $\Lambda$CDM cosmology provided by
\citet{Planck2015} in which the universe is flat with $H_0 = 67.7$\,km\,s$^{-1}$
and $\Omega_m = 0.307$ and report all magnitudes in the AB system.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{Imaging of the field of GRB~111117A with the X-shooter slit overlaid. Only one slit is shown, despite four epochs of spectroscopic observations because of the similarity in position angle. The image is the FORS2 $R$-band image and%, for which the photometry is shown in Fig. \ref{fig:SED}. 
	the blue asterisk indicates the GRB position as derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. The errorcircle of the \emph{Chandra}-position is too small to be visible in the image.}
	\label{fig:spec_setup}
\end{figure}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, in four seperate epochs. The
burst was followed up 38 hours after the Burst Alert Telescope (BAT) trigger
under ESO programme 088.A-0051 (PI: Fynbo) and again later using a different ESO
program 091.D-0904 (PI: Hjorth). X-shooter can cover the wavelength range from
3000~\AA~to 24\,800~\AA~(21\,000~\AA~when the $K$-band blocking filter is used) across
three spectroscopic arms, simultaneously illuminated through the use of
dichroics. The bias-correction, flat-fielding, order tracing, wavelength
calibration, rectification, and flux calibration is then carried out using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010}. 
%Because the spectra are curved across each detector, a rectification algorithm
%is employed which introduces correlations between neighboring pixels. We select
%a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the
%degree of correlation while conserving the maximal resolution.
The observations are combined and extracted using scripts described in Selsing
et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The
signal-to-noise of the continuum in the near-infrared arm is too low to use the
optimal extraction algorithm \citep{Horne1986} and therefore the extraction is
carried out with a simple aperture.

An overview of the spectroscopic observations is given in Table~\ref{tab:spec_overview}, and the position of the slit on the target is shown in Fig.~\ref{fig:spec_setup}.

@@

We determine a redshift of $z = 2.211$ from the simultaneous detection of
emission lines interpreted as \lya, \oii, \hb, \oiii$\lambda$5007, and \ha, with
\hb~detected at low significance ($\sim 3 \sigma$). \ha~is only visible in the
first epoch, due to the $K$-band blocking filter used for the remainder
observations. The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and
show significant velocity structure along the slit. A drop in the continuum to
the blue of the \lya~line further supports the inferred redshift.

@@

% TK: Some rephrasing below.

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of a 
galaxy \citep{Kennicutt1998}. At the redshift of the GRB host, \ha~is observed at 
21\,000~\AA~where the night sky is very bright. In addition, several bright sky-lines 
intersect the line, making an accurate estimate of the \ha-flux difficult. 
We obtain a limit on the SFR by integrating the part of \ha~free of contamination and 
correcting for the missing flux using the line shape. After converting the 
\citet{Kennicutt1998} relation to a \citet{Chabrier2003} initial mass function using
\citet{Madau2014}, we derive a limit of $SFR > 7 M_\odot$~yr$^{-1}$. 
From the SED-fit (Sect. \ref{SED}), and the detection of \lya, the host is
constrained to contain very little or no dust, which is why we do not apply 
a dust-correction to the measured \ha~flux here.

@@

In addition to the spectroscopy presented above, we imaged the field of GRB~111117A 
in multiple broad-band filters using the VLT equipped with FORS2 ($gRIz$ filters) and HAWK-I
($JHK_{\mathrm{s}}$ filters), long after the burst has faded. These new data are
complemented by a re-analysis of some of the imagery used in
\citet{Margutti2012} and \citet{Sakamoto2013} that are available to us here (GTC
$gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in Table~\ref{tab:phot_overview}.

% TK: I shortened a bit here
All data were reduced, analyzed and fitted in a similar manner as described in detail in
\citet{Kruhler2011a} and, more recently, in \citet{Schulze2016}. Very briefly, we use our own
\texttt{python} and IRAF routines to perform a standard reduction which includes
bias/flat-field correction, de-fringing (if necessary), sky-subtraction, and
stacking of individual images. The photometry of the host was tied against magnitudes 
of field stars from the SDSS and 2MASS catalogs in the case of $grizJHK_{\mathrm{s}}$ filters. 
For the $R$ and $I$-band photometry we used the color transformations of
Lupton\footnote{\url{https://www.sdss3.org/dr8/algorithms/sdssUBVRITransform.php}}. 
We convert all magnitudes into the AB system if necessary, and correct for the Galactic 
foreground of $E_{B-V}=0.027~\mathrm{mag}$.

The multi-color spectral energy distribution (SED) is fit by
\citet{Bruzual2003} single stellar population models based on a
\citet{Chabrier2003} initial mass function in \emph{LePhare} \citep{Ilbert2006},
where the redshift is fixed to the spectroscopic value of $z=2.211$. The best
model is obtained with an unreddened galaxy template, and returns physical
parameters of absolute magnitude ($M_B=-22.0\pm0.1$\,mag), stellar mass
($\log(M_{\star}/M_\odot) = 9.9\pm0.2$), stellar population age ($\tau =
0.5_{-0.3}^{+0.5}$ Gyr) and star-formation rate
($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). %We show the SED
% fit in Fig. \ref{fig:SED}.

Noteworthy is the discrepancy of our new VLT/FORS2 photometry and the
re-analysis of Gemini data to the $z$-band measurements of \citet{Margutti2012} and
\citet{Sakamoto2013}. Both of these authors report $z$-band photometry that is brighter 
by 0.8~mag to 1.0 mag than our value, whereas data taken in bluer filters are in excellent agreement. In fact, the large $i-z$ color is mistakenly interpreted as a 4000\,\AA\,break
driving the galaxy photometric redshift of the earlier works. Using
the revised photometry from Table~\ref{tab:phot_overview}, the photometric redshift of the
galaxy is $z_{\mathrm{phot}}=2.04_{-0.21}^{+0.19}$, consistent with the
spectroscopic value at the 1~$\sigma$ confidence level.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{X-ray temporal and spectral analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We retrieved the automated data products provided by the \textit{Swift}-XRT GRB
repository\footnote{\url{http://www.swift.ac.uk/xrt\_products/00507901}} \citep{Evans2009}. 
% TK: Fixed the link
The X-ray afterglow light curve
can be fit with a single power-law decay with index
$\alpha=1.27_{-0.10}^{+0.12}$. We performed a time-integrated spectral analysis
using data in photon counting (PC) mode in the widest time epoch where the
$0.3-1.5\,\mathrm{keV}$ to $1.5-10\,\mathrm{keV}$ hardness ratio is constant
(namely, from $t-T_0 = 205$~s to $t-T_0 = 203.5$~ks, for a total of 29.1~ks of
data) to prevent spectral changes that can affect the X-ray column
density determination. The obtained spectrum is well described by an absorbed
power-law model. The best-fit spectral parameters are a photon index of $2.1 \pm
0.4$ and an intrinsic hydrogen column density $N_{\mathrm{H}}$ of $2.4_{-1.6}^{+2.4} \times 10^{22}$~cm$^{-2}$ ($z=2.211$), assuming a Galactic $N_{\mathrm{H}}$ in the burst direction of $4.1 \times 10^{20}$~cm$^{-2}$.

@@

As already pointed out \citep{Margutti2012, Sakamoto2013}, GRB~111117A securely 
belongs the short class of GRBs. Because the observed classification
indicators, $T_{90}$ and hardness ratio, do not depend strongly on redshift
\citep{Littlejohns2013a}, the updated redshift does not change this designation
significantly. The intrinsic luminosity increases, as visible in the X-ray light curve
in Fig. \ref{fig:sxray_lightcurve}, but it still is sub-luminous compared to
the majority of long GRBs. \citet{Bromberg2013} investigated the degree to which the
long and short population distributions overlap and quantified the certainty in
the class membership. GRB~111117A has $96_{-5}^{+3}$ percent probability of
being short. Compared to the other two short GRB candidates at high redshift, GRB060121
\citep{DeUgartePostigo2006, Levan2006} at $1.7 \lesssim z \lesssim 4.5$
($17_{-15}^{+14}$ per cent) and GRB090426 \citep{Antonelli2009, Levesque2010,
	Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$ per cent), the certainty in class
membership for GRB~111117A is much higher.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/XLC_111117A_rest.pdf}
	\caption{Restframe XRT lightcurve of GRB~111117A, compared to the general population of XRT lightcurves of GRBs. The grey shared region is a compilation of long GRB lightcurves where the color represents density and the light blue is other short GRB lightcurves for which redshifts has been determined. Despite the remarkably high redshift, the luminosity is comparable to the bulk of the short burst population and still subluminous compared to the lGRB population.}
	\label{fig:sxray_lightcurve}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Restframe $N_\mathrm{H}$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We plot the recalculated $N_\mathrm{H}$ in Fig. \ref{fig:NH_z} where we compare with the
distributions of complete samples of both long and short GRBs. The long sample
is from \citet{Arcodia2016} and the short sample is from \citet{DAvanzo2014a}.
17 of the 99 long bursts do not have redshifts and likewise 5 out of 16 for
the short sample. Bursts without redshifts have been excluded for both groups.
GRB~111117A occupies a unique position in Fig. \ref{fig:NH_z} with the highest
$N_H$ of all short burst. The short sample, excluding GRB~111117A, is located at
low redshifts ($z < 1$) and is found to populate a similar column density
environment to long GRBs at similar redshifts \citep{DAvanzo2014a}. The inferred
hydrogen column for GRB~111117A seems to follow the trend with increasing $N_H$
as a function of redshift as found for the long GRB afterglows \citep{Arcodia2016}.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_z.pdf}
	\caption{Rest frame, X-ray derived hydrogen column densities for GRB~111117A compared to complete samples of both long and short populations. In red we show the sample of lGRBs from \citet{Arcodia2016}. The detections are replaced with contours for clarity and the limits are shown with arrows. In blue is the sample of sGRBs from \citet{DAvanzo2014a}. GRB~111117A is an outlier from the short sample, but seems to follow the distribution of long GRBs.}
	\label{fig:NH_z}
\end{figure}

@@

%From the clear host association, GRB~111117A does not belong to the hostless class of GRBs \citep{Berger2010a} and because the host exhibits clear emission lines indicative of a population of relatively young stars, 
As the majority of short GRBs \citep{Fong2013b}, the host of GRB~111117A is a
late-type galaxy and is entirely consistent in terms of stellar mass and stellar
age with the general population of short GRB hosts ($\left\langle M _*
\right\rangle = 10^{10.1} M_{\odot}$ and $\left\langle \tau _* \right\rangle =
0.3 $~Gyr) \citet{Leibler2010}. The SFR, on the other hand, is $\sim$1 order of
magnitude higher than the typical SFR for short GRB hosts galaxies
\citep{Berger2014} and more similar to the SFRs found in the hosts of long GRBs
at a corresponding redshift \citep{Kruhler2015}. Only two hosts in the sample of
short GRBs compiled in \citet{Berger2014} have a more vigorous star formation,
meaning that it is in the upper end of the star formation distribution,
while not a complete outlier. This is likely a selection effect, because a less
star-forming galaxy would exhibit weaker emission lines, thus making the
redshift harder to determine. Being a late-type host, both the stellar mass and
sSFR are entirely consistent with the expected host of
sGRBs \citep{Behroozi2014}.

The detection of \lya~is consistent with the SED-inferred absence of dust, despite
the moderate stellar age and the high X-ray derived hydrogen column density
which would suggest the opposite. The centroid of the Ly$\alpha$ emission is
found to be redshifted by $\sim$~220~km~s$^{-1}$ with respect to systemic, which
is similar to what is found for long GRB hosts \citep{Milvang-Jensen2012a} where
the outflow is attributed to star formation.
